Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes

### 2024-10-07 ms/ms.Rmd
ms: models rf + rk

>>>>

We used Random Forests (RF) to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in digital soil mapping applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.

RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004].
If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that simple RF rather regression kriging was justified at both sites.

### 2024-10-07 ms/ms.tex
ms: models rf + rk

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative the predictor space of the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.

@@

We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to greatly improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, newmanAssessingSpatiallyHeterogeneous2023}.
We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.

@@

>>>>

At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.

@@

We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway \citep{maxwellLandsurfaceParametersSpatial2022}.
We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.

@@

We used Random Forests (RF) to predict peat depth at both sites.
RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in digital soil mapping applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.

RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004}.
If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that simple RF rather regression kriging was justified at both sites.
